On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-25 tex/draft/draft4.tex
summary edits

%In this paper we study the formation of giant planets through core accretion.

In this paper we build evolutionary models of giant planets forming in a protoplanetary disk. We consider atmosphere growth around a core of fixed mass, and determine the minimum core mass for which unstable atmosphere collapse occurs before the dissipation of the protoplanetary disk, at various locations in the disk. For a standard ISM power-law opacity and a polytropic EOS for the nebular gas, the minimum core mass is $\sim 8 M_{\oplus}$ at 5 AU and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU. A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low.  Our results are applicable in the outer regions of the disk ($\gtrsim 5$ AU), where planets can undergo runaway gas accretion before opening a gap in the disk. If follows that core accretion at wide separations may result in massive gas giants, which could explain the formation of some of the directly imaged giant planets (\citealt{marois08}; \citealt{kalas08}; \citealt{lagrange10}).

Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies. A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime. We note that our model is a limiting case of core accretion, in which planetesimal accretion is negligible and the evolution of the gaseous envelope is dominated by KH contraction. We are thus able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool. Our results are not inconsistent with the existence of ice giants at large distances (such as Uranus and Neptune in our own Solar System), as ongoing accretion of solids can stall atmospheric growth and prevent the accumulation of a massive envelope before disk dissipation.

% In this paper we study the formation of giant planet atmospheres under the assumption of a low planetesimal accretion regime in which KH contraction dominates the luminosity evolution of the atmosphere. We consider atmosphere growth around a core of fixed mass, and we build quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk at the Hill radius. In our static models, we use an ideal gas polytropic EOS, and assume a standard ISM power-law opacity for dust grains. We derive a cooling model to connect series of quasti-static atmospheres and obtain an evolutionary history of the envelope. We define the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, for which $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$. From this we define as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk. We calculate the critical core mass for a variety of disk conditions, nebular gas compositions and opacities. The critical core mass is $\sim 8 M_{\oplus}$ at 5 AU in our fiducial disk and decreases to $\sim 3.5 M_{\oplus}$ at 100 AU. A larger mean molecular weight of the nebular gas increases the critical core mass, while a factor of ten reduction in opacity results in a critical core mass twice as low. 
% 
%Our results are lower than the typically quoted critical core mass of 10 $M_{\oplus}$ in static studies. A giant planet can thus form faster if the core grows first, then accretion of solids slows down and a massive atmosphere accumulates, rather than if the core and envelope grow simultaneously in a high planetesimal accretion rate regime. Moreover, we are able to place robust absolute lower limits on the core mass needed to form a giant planet, since additional core growth and heating limit the ability of the atmosphere to radiate energy and cool.

 %We also find that the critical core mass to form a giant planet within the life time of the disk (of about 3 Myrs) is smaller than the typical 10 $M_{\oplus}$ value yielded by studies that assume that the atmosphere evolution is dominated by the luminosity due to planetesimal accretion. As such, it is faster to form a planet by growing the core first in a fast planetesimal accretion regime (e.g., the core forms in the inner disk, then migrates outwards), then significantly reduce planetesimal accretion and allow a massive atmosphere to accumulate.

>>>>

We study the formation of giant planets by the core accretion mechanism.  Our models start with a solid core that is embedded in a gas disk and no longer accreting solids.  We determine -- as a function of disk location, core mass, and the atmosphere's mean molecular weight and opacity -- whether runaway atmospheric growth can occur within a typical disk lifetime of 3 Myr.  By neglecting the accretion luminosity of planetesimals and smaller solids, we obtain the fastest allowed rate of gas accretion.  Of relevance to direct maging surveys, we address core accretion in the outer disk.  Our model approximations, including spherical accretion and low mass radiative exteriors with opacities dominated by dust, are tuned to conditions in the outer disk.   Our main findings are as follows:

\begin{enumerate}
\item The minimum or critical core mass, $M\crit$, for giant planet formation declines with distance in standard protoplanetary disk models.  For our reference case, the critical mass is $\sim 8.5 M_{\oplus}$ at $a = 5$ AU, decreasing to $\sim 3.5 M_{\oplus}$ at $a =100$ AU.  This decline roughly follows $M\crit \propto a^{-0.3}$.

\item The drop in disk temperature with radial distance explains the decrease in critical core masses.  The lower pressures and densities in the outer disk only weakly suppress atmospheric growth.

\item Reducing dust opacities by a factor of 10 reduces critical core masses by a factor of $\sim 3$.  This reduction is somewhat stronger (weaker) at larger (smaller) separations from the star.

\item A larger mean molecular weight reduces critical core masses, in agreement with \citet{HorIko11}.  If enrichment in heavy elements correlates with increased dust opacity, then the stronger opacity effect will dominate, increasing $M\crit$.

\item Runaway growth begins roughly at the crossover mass, when atmosphere and core masses are equal, in agreement with previous work \citep{pollack96}.  For larger core masses, runway growth begins at larger ratios of atmosphere to core mass.

\item Self-gravity affects atmospheric evolution before crossover.  Significant self-gravitational corrections appear when the atmosphere is only $\sim 20 \%$ as massive as the core.

\end{enumerate}

Rapid gas accretion onto low mass cores could explain the origin of distant directly imaged giant planets \citep{marois08, lagrange10}.  However our model does address the details of how solid cores grow, and many possibilities exist.  For a giant planet to form with a core near the minimum mass, the core's growth must  first be rapid and then slow significantly, as in phases I and II, respectively, of \cite{pollack96}.

Initial core growth must be fast, compared to the disk lifetime, to get a sufficiently massive core.  Subsequent slowing of core growth is required to allow the atmosphere to cool and contract.  To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.  Specifically a core mass doubling timescale of $\sim 400$ Myr would provide enough heating to stall atmospheric cooling and growth.

Thus for giant planets to form around around cores near $M\crit$, core growth must stall quite severely.   The accretion and scattering of local planetesimals, combined with the inward migration of smaller solids past the core, offers a plausible means to sufficiently isolate the core.  On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk.  In these cases, relatively high mass cores would remain as ice giants, akin to Uranus and Neptune.   Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.
